---
title: "Procrustes analysis via diffeomorphic image matching"
author: "BB Avants, JT Duda, NJ Tustison"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

The singular value decomposition (and its close relative, principal component analysis) is perhaps the most widely used dimensionality reduction method in statistics and machine learning.  Both efficient implementations and mature approaches to visualization make the method accessible and interpretable in a broad range of applications.  However, several limitations exist.  Primarily, the basis produced by SVD is severely limited by the number of samples.  Relatedly, the basis is not sparse in the feature space.  Both of these concerns may be addressed by methods such as sparse PCA, ICA or NMF.  However, these methods introduce additional assumptions that may not be easy to check (e.g. FIXME), additional parameters that may not be intuitive to select (e.g. weights for sparseness penalties) and the algorithms may not be implemented efficiently enough to be effective in datasets that may be large in $n$ and/or $p$.

# Methods

# Procedure 

Load the deformation field dataset and the template that defines the coordinate
system in which the deformations are defined. This data is computed in the example:
[ANTs template building example](https://github.com/ntustison/TemplateBuildingExample).

```{r spcaexample1}
library( ANTsR )
# should set this directory manually!
bd=path.expand( "~/data/antsExamples/TemplateBuildingExample/Faces/" )
if ( ! dir.exists( bd ) )
  stop("set base directory to point to template face example output.")
defs = Sys.glob( paste(bd,'TemplateFaces/T_face*[0-9]Warp.nii.gz',sep='') )
defsi = Sys.glob( paste(bd,'TemplateFaces/T_face*[0-9]InverseWarp.nii.gz',sep='') )
affs = Sys.glob( paste(bd,'TemplateFaces/T_face*GenericAffine.mat',sep='') )
faces = Sys.glob( paste(bd,'fa*.nii.gz',sep='') )
imgs = Sys.glob( paste(bd,'TemplateFaces/T_template0face*WarpedToTemplate.nii.gz',sep='') )
template = antsImageRead( paste(bd,'TemplateFaces/T_template0.nii.gz',sep='') )
msk = getMask( template ) %>% iMath( "MD", 8 )
jlist = list() 
for ( k in 1:length( defs ) )
  jlist[[ k ]] = createJacobianDeterminantImage( template, defs[[ k ]] )
imat = imageListToMatrix( jlist, msk )
# msk = antsImageRead( "eye.nii.gz" )
wlist = list( )
for ( i in 1:length( defs ) ) wlist[[ i ]] = antsImageRead( defs[i] )
```

Compute the PCA basis.

```{r computer}
if ( !exists("dpca") )
####
#  dpca = multichannelPCA( wlist, msk, pcaOption='randPCA', verbose=FALSE )
# dpca = multichannelPCA( wlist, msk, k=10, pcaOption='eanat', verbose=TRUE )
# dpca = multichannelPCA( wlist, msk, k=50, pcaOption=2000, verbose=FALSE, auxiliaryModality = imat )
 dpca = multichannelPCA( wlist, msk, k=8, pcaOption='fastICA', verbose=TRUE )
####
```

Show the shape change and its magnitude / locality.

```{r viewer}
mxk=ncol( dpca$pca$v )
for ( ww in 1:mxk )
  {
  print( ww )
  myw = dpca$pcaWarps[[ww]] * 0.25 / max( abs( dpca$pcaWarps[[ww]]  ) )
  myw = smoothImage( myw, antsGetSpacing( template ) * 5.0 )
  warpTx = antsrTransformFromDisplacementField(  myw  )
  # compose several times to get a visual effect
  wtxlong = list( ) ; for ( i in 1:20 ) wtxlong[[i]]=warpTx
  warped = applyAntsrTransform( wtxlong, data = template,
    reference = template )
  for ( kk in 1:1 ) {
    plot( template )
    plot( warped )
    }
  # look at the magnitude ...
  splt = splitChannels( dpca$pcaWarps[[ww]]  )
  mgntd = iMath( abs( splt[[1]] ) + abs( splt[[2]] ), "Normalize" )
  plot( template, mgntd, window.overlay=c(0.1,1) )
  }
```

Create a random face.

```{r randspcaresult}
# now make a "random" face
combvec = dpca$pcaWarps[[1]]
for ( i in 2:mxk )
  combvec = combvec + dpca$pcaWarps[[i]] * rnorm(1, sd=5)
combvec = combvec * 0.25 / max( abs( combvec  ) )
warpTx = antsrTransformFromDisplacementField( combvec  )
# compose several times to get a visual effect
wtxlong = list( ) ; for ( i in 1:12 ) wtxlong[[i]]=warpTx
warped = applyAntsrTransform( wtxlong, data = template,
  reference = template )
for ( kk in 1:1 ) {
  plot( template )
  plot( warped )
  }
```



The previous examples show how we can use an unbiased decomposition 
of the data in order to understand shape variation.  However, this 
full decomposition may obscure details that we may be able to see if
we focus on specific landmarks, as in traditional geometric morphometry [ref].


```{r geomorph0}
library( ANTsR )
library( geomorph )
# read the landmarks
template = antsImageRead( path.expand( "~/data/antsExamples/isa/data/template.nii.gz" ) )
lm = antsImageRead( path.expand( "~/data/antsExamples/isa/data/landmarks.nii.gz" ) )
plot( template, lm )
mydim = lm@dimension
```


Map all of these landmarks to the subject space to treat them as landmarks.

```{r geomorph}
lmCents = getCentroids( lm, clustparam = 3 )[,1:mydim]
plot( lmCents[,1], lmCents[,2]*-1, main='template and subject points')
pointmat = matrix( nrow = length( defs ), ncol = length( lmCents ) )
mycolors = palette( rainbow( length( defs ) ) )
for ( k in 1:length( defs ) )
  {
  loctx = c( defs[k], affs[k] )
  loctxi = c( affs[k], defsi[k] )
  facek = antsImageRead( faces[ k ] )
  lmWarpImage = antsApplyTransforms( facek, lm, transformlist=loctxi, 
    interpolator = 'nearest', whichtoinvert = c(T,F) )
  mycents = getCentroids( lmWarpImage, clustparam = 3 )[1:p,1:mydim]
  pointmat[ k ,  ] = as.numeric( mycents - lmCents  )
  points( mycents[,1], mycents[,2]*(-1), col=mycolors[ k ] )
  plot( facek, lmWarpImage )
#  Sys.sleep( 5 )
  }
```

Run pca on the point deformation matrix.

```{r mypca,eval=TRUE}
mypca = svd( pointmat )
plot( ts( mypca$d / sum( mypca$d ) ) )
# convert back to a deformation
defmat = matrix( t(mypca$v), nrow=length(defs), ncol=length( lmCents ) )
for ( k in 1:nrow(defmat) )
  {
  def1 = lmCents + defmat[,k] * 4.0
  temp = rbind( lmCents, def1 )
  plot( temp[,1], temp[,2]*-1, main='template and Kth deformation', cex=1.5, lwd=3 )
  points( def1[,1], def1[,2]*-1, col=mycolors[ k ], cex=1.5, lwd=3 )
  }
```


```{r gpa,eval=FALSE,echo=FALSE}
# Run generalized procrustes analysis.
Y.gpa <- gpagen( myarr, PrinAxes = FALSE, max.iter=1  )
temp = Y.gpa$consensus
plot( temp[,1], temp[,2], main='template and subject points')
```


```{r gpaviz1,eval=FALSE,echo=FALSE}
#GPA-alignment
plotTangentSpace( Y.gpa$coords )
```


```{r gpaviz2,eval=FALSE,echo=FALSE}
ref<-mshape( Y.gpa$coords )
plotRefToTarget( ref, ref )
k=5
plotRefToTarget( ref, Y.gpa$coords[,,k])
GP3 <- gridPar(pt.bg = "blue", pt.size = 1.5, tar.pt.bg = "orange", tar.pt.size = 1)
plotRefToTarget(ref,Y.gpa$coords[,,k], gridPars=GP3, mag=3, method="vector")
```
